Extended Data

### 2022-03-13 doc/manuscript/Dow_manuscript_Nature.Rmd
update word count
#125
wordsmithing - results

<!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text). We're currently at 2763 words and 1 modest + 2 large display items. Articles typically have 30-50 references (currently at 53)-->

@@

Both MODIS-derived canopy foliage phenology and dendrometer band measurements of stem growth timing generally shifted backwards as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem growth timing were found using $T_{max}$ during a critical temperature window (CTW). 
CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 to mean $DOY_{25}$ for each xylem architecture-site combination (`r EDF_climwin`).
The CTW was defined as the week(s) which had the strongest correlation with $DOY_{25}$.

![**Figure 2 | Responses of foliage phenology (a,b) and stem growth timing (c,d) to spring temperatures at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each ForestGEO plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down). Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring porous and diffuse porous trees at SCBI (2011-2020) and Harvard Forest (1999-2003). Mean $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`. Mean temperature was calculated for each xylem architecture/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red. Leaf phenology years are colored according to temperature in the critical $T_{max}$ window which had the best correlation with timing of greenup (ring porous window at HF and diffuse porous at SCBI).](tables_figures/DOYtiming_allyears.png){ width=25% }

For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of seasonal milestones.
Consistent with findings from previous studies[<!--@parmesan_globally_2003-->;<!--@ibanez_forecasting_2010-->;<!--@friedl_tale_2014-->], leaf phenological milestones advanced at both sites (Fig. `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).<!--Note to Cam: double check that these were updated-->
Similarly, with the exception of ring porous $DOY_{75}$ at SCBI, the stem growth milestones $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ decreased with mean $T_{max}$ during the CTW (Figs. `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ advanced 0-1.7 days/ $^\circ$C for ring porous species and 2.8-2.9 days/ $^\circ$C for diffuse porous species at SCBI, and 10.3-12.3 days/ $^\circ$C for ring porous species and 0.9-4.2 days/ $^\circ$C for diffuse porous species at Harvard Forest (`r EDT_phenology_parameters`).

Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer temperatures during the CTW compared to those with cooler temperatures (Fig. `r DOY_timing`a-b), there was no consistent lengthening of $L_{PGS}$ (Fig. `r schematic_summary`, `r EDF_pheno_all`).<!--Note to Cam: double check this-->

In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were weak and inconsistent (Figs. `r schematic_summary`, `r EDF_pheno_all`). 
Specifically, $g_{max}$, which occurred very close to $DOY_{50}$, displayed extremely small negative changes (HF) or changes in opposite directions (SCBI) in relationship to mean $T_{max}$ during the CTW for ring- and diffuse- porous species.
$\Delta DBH$ displayed a weak positive relationship (+0.029-0.08 mm/ $^\circ$C) with mean $T_{max}$ during the CTW at SCBI and a weak negative relationship (-0.2-0.03mm/ $^\circ$C) at HF (`r EDF_pheno_Tmax_climwin`).

>>>>

<!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text). We're currently at 2641 words and 1 modest + 2 large display items. Articles typically have 30-50 references (currently at 53)-->

@@

Both canopy foliage phenology and stem growth timing shifted earlier as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem growth timing were found using $T_{max}$ during a critical temperature window (CTW). 
CTW was identified by measuring the correlation between weekly $T_{max}$ and $DOY_{25}$ for all combinations of consecutive weeks from January 1 to mean $DOY_{25}$ for each xylem architecture-site combination (`r EDF_climwin`).
The CTW was defined as the weeks for which this correlation was strongest, and mean $T_{max}$ during this period (CTW $T_{max}$) used as the independet variable.

![**Figure 2 | Responses of foliage phenology (a,b) and stem growth timing (c,d) to spring temperatures at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each ForestGEO plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down). Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring porous and diffuse porous trees at SCBI (2011-2020) and Harvard Forest (1999-2003). Mean $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`. Mean maximum temperature ($T_{max}$) was calculated for each xylem architecture/site combination over the respective critical temperature window (CTW), then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red. Leaf phenology years are colored according to the CTW $T_{max}$ with the best correlation with timing of greenup (diffuse porous window at SCBI and ring porous window at Harvard Forest).](tables_figures/DOYtiming_allyears.png){ width=25% }

For ring- and diffuse- porous species at both sites, warmer CTW $T_{max}$ resulted in earlier achievement of seasonal milestones.
Consistent with findings from previous studies[<!--@parmesan_globally_2003-->;<!--@ibanez_forecasting_2010-->;<!--@friedl_tale_2014-->], leaf phenological milestones advanced at both sites (Fig. `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).<!--Note to Cam: double check that these were updated-->
Similarly, with the exception of ring porous $DOY_{75}$ at SCBI, the stem growth milestones $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ decreased with CTW $T_{max}$ (Figs. `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ advanced 0-1.7 days/ $^\circ$C for ring porous species and 2.8-2.9 days/ $^\circ$C for diffuse porous species at SCBI, and 10.3-12.3 days/ $^\circ$C for ring porous species and 0.9-4.2 days/ $^\circ$C for diffuse porous species at Harvard Forest (`r EDT_phenology_parameters`).

Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased with CTW $T_{max}$ (Fig. `r DOY_timing`a-b)<!--specify which species group used?-->, there was no consistent lengthening of $L_{PGS}$ (Fig. `r schematic_summary`, `r EDF_pheno_all`).<!--Note to Cam: double check this-->

In contrast to the pronounced effects of CTW $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were weak and inconsistent (Figs. `r schematic_summary`, `r EDF_pheno_all`). 
Specifically, $g_{max}$, which occurred very close to $DOY_{50}$, displayed extremely small negative changes (HF) or changes in opposite directions (SCBI) in relationship to CTW $T_{max}$ for ring- and diffuse- porous species.
$\Delta DBH$ displayed a weak positive relationship (+0.029-0.08 mm/ $^\circ$C) with CTW $T_{max}$ at SCBI and a weak negative relationship (-0.2-0.03mm/ $^\circ$C) at HF (`r EDF_pheno_Tmax_climwin`).
